---
title: '5th Protocol: Precipitation'
author: "Cheyenne Rueda and Simone Massaro"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document:
    toc: true
    includes:
      in_header: "../preamble.tex"
header-includes:
  - \fancyhead[L]{Shortwave radiation}
---

\newpage
\newpage

# Precipitation

## Motivation

Precipitation is a very important process in climatology or weather meteorology. Precipitation in relation with water, occurs when atmospheric conditions become saturated enhancing 100% of relative humidity and thus, water condensates and later, it precipitates. Besides, precipitation is a key process part of the water cycle. Its function is to provide with water the natural ecosystem in the planet. Water is the first component needed by plants, to allow the processes of photosynthesis and transpiration.


## Background

The precipitation is usually in form of liquid water, but can also be solid in case of snow or hail. In the previous section, we referred to water condensation. Although this is important part, first needs to be a cloud formation. This takes place by the presence of a warm air flow, thus its cooler surrounding will rise due to a lower density. Together with an increase in height, the air parcel expands and cools down with the dry adiabatic lapse rate. Then, condensation occurs when air temperature and dew point temperature. During the condensation latent heat is released.

It is measured as a volume of water per unit area ($m^3/m^2$). The standard unit is the millimeters (mm) of rain, which corresponds to a liter per squared meter.
The precipitation intensity is measured in millimeter per hour ($mm/h$)

## Sensors and measuring principle

The precipitation sensors, pluviometers, measure the amount of rainfall on a know area.
The most common ones uses a **tipping bucket** that when gets filled with water it switch position and empty itself. By counting the number of swings of the bucket it is possible to calculate the amount of precipitation.
This sensors are the most common one, as they are simple and relatively reliable, however they can produce inaccurate measurements if the amount of precipitation is very low that the buckets doesn't fill or when there is an high rain intensity that the bucket cannot fill fast enough.

To overcome this limitation there are **weighting** pluviometers, which have a very sensitive scale under a bucket that collects all the precipitations.

A completely different type of sensors are the **laser beam** pluviometers, that can estimate the number, size and speed of water droplet and therefore produce an accurate measurement of rain.


Precipitation measurements present additional challenges, in presence of strong winds there is turbulence around the instrument that reduce the amount of rain collected. To mitigate this there are wind shield that can be installed around pluviometers.
Moreover precipitation can be solid, hence heating may be required to properly measure it. Finally precipitation has an high spatial heterogeneity, which requires the use of several sensors.


\newpage

## Analysis
```{r, include = F}
library(tidyverse)
library(patchwork)
library(here)
```
```{r, message=F}
prec <- read_csv(here("Data_lectures/5_Precipitation/P_4sites.csv"))
et <- read_csv(here("Data_lectures/5_Precipitation/ET_4sites.csv"))
sites <- read_csv(here("5_precipitation/station_data.csv")) %>%
  rename(site=`Site-abb`, full_name = Site)
```
```{r}
prec_avg <- prec %>%
  select(-Date) %>%
  gather("site", "prec") %>%
    group_by(site) %>%
  summarise(prec=mean(prec))

et_avg <- et %>%
  select(-Date) %>%
  gather("site", "et") %>%
    group_by(site) %>%
  summarise(et=mean(et))

sites_avg <- sites %>%
  inner_join(et_avg, by="site") %>%
  inner_join(prec_avg, by="site")
```

### ET and precipitation at different latitudes

The precipitation and the Evapotranspiration (ET)

```{r}
prec_t <- prec %>%
  gather("site", "prec", -Date)

et_t <- et %>%
  gather("site", "et", -Date) 

prec_et <- left_join(prec_t, et_t, by=c("Date", "site"))

sites <- sites %>%
  inner_join(prec_et, by="site") 
```




```{r prec, fig.cap="Precipitation"}
(prec_box <-  ggplot(sites, aes(Lat, prec, colour=site)) +
  geom_boxplot() +
   labs(title="Cumlative precipitation at different latitudes",
        y="Total yearly precipitation (mm)",
        x="Latitude (deg)"))
```

\newpage

```{r et, fig.cap="Cumlative evapotranspiration at different latitudes"}
(et_box <- ggplot(sites, aes(Lat, et, colour=site)) +
  geom_boxplot() +
    labs(y="Total evapotranspiratio [mm]",
        x="Latitude [deg]")) +
  scale_color_colorblind()
```

### Evapotranspiration index

Precipitation is used as a driver/source for evapotranspiration. The ratio between
evapotranspiration and precipitation (evapotranspiration index = ET_idx) is a very useful indicator
how precipitation is used by plants during photosynthesis. Calculate the ET_idx for all sites and
visualise it. What surprised you and how do the ET_idx differs across ecosystems? Discuss.


```{r et-index, fig.cap="Evapotranspiration index"}
sites$et_idx <- sites$et/sites$prec

sites %>% 
ggplot(aes(site, et_idx, colour = site))+
         geom_boxplot()+
         labs( x="Different site", y="Evapotranspiration index") +
  scale_color_colorblind()
        
```

As we can see,evapotranspiration index is the fraction between precipitation and what is coming back to atmosphere through evapotranspiration. 

what is intercepted by cannopy and what is being lose by soil hydrology processes. Other variables such as wind speed, surface humidity and solar radiation are not being taken into account in the calculation of this index, which probably are the ones causing differences on what it could have been expected at the beginning for some of the places. 

we can see a higher evapotranspiration index in places where there is plenty precipitation and evapotranspiration rate by canopy. Hainich and Hyytiala are represented as the higher, this can be due to their higher latitudes providing this ecosystems with more precipitation rate. Guyaflux is at lower latitude and thus more close to the equator. This means a more tropical weather where high amount of rain and sun is normally occurring. Collelongo is place in a drier climate with high radiation but less annual precipitation rates. 





