---
title: '5th Protocol: Precipitation'
author: "Cheyenne Rueda and Simone Massaro"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document:
    toc: true
    includes:
      in_header: "../preamble.tex"
header-includes:
  - \fancyhead[L]{Shortwave radiation}
---
# Precipitation

## Motivation

## Background

## Sensors and measuring principle

## Analysis
```{r, include = F}
library(tidyverse)
library(patchwork)
library(here)
```
```{r, message=F}
prec <- read_csv(here("Data_lectures/5_Precipitation/P_4sites.csv"))
et <- read_csv(here("Data_lectures/5_Precipitation/ET_4sites.csv"))
sites <- read_csv(here("5_precipitation/station_data.csv")) %>%
  rename(site=`Site-abb`, full_name = Site)
```
```{r}
prec_avg <- prec %>%
  select(-Date) %>%
  gather("site", "prec") %>%
    group_by(site) %>%
  summarise(prec=mean(prec))

et_avg <- et %>%
  select(-Date) %>%
  gather("site", "et") %>%
    group_by(site) %>%
  summarise(et=mean(et))

sites_avg <- sites %>%
  inner_join(et_avg, by="site") %>%
  inner_join(prec_avg, by="site")
```

## ET and precipitation at different latitudes

*Visualise the annual sums of evapotranspiration and precipitation according to its geographic
latitude (boxplot?). How and why does precipitation and evapotranspiration varies? Which impact
has the underlying ecosystem?*

```{r}
# this t makes no sense as name TODO chanve this
prec_t <- prec %>%
  gather("site", "prec", -Date)

et_t <- et %>%
  gather("site", "et", -Date) 

prec_et <- left_join(prec_t, et_t, by=c("Date", "site"))

sites <- sites %>%
  inner_join(prec_et, by="site") 
```

```{r}
sites
```


```{r}
(prec_box <-  ggplot(sites, aes(Lat, prec, colour=site)) +
  geom_boxplot() +
   labs(title="Cumlative precipitation at different latitudes",
        y="Total yearly precipitation (mm)",
        x="Latitude (deg)"))
```
```{r}
(et_box <- ggplot(sites, aes(Lat, et, colour=site)) +
  geom_boxplot() +
    labs(title="Cumlative evapotranspiration at different latitudes",
        y="Total yearly evapotranspiratio (mm)",
        x="Latitude (deg)"))
```
## Evapotranspiration index

*Precipitation is used as a driver/source for evapotranspiration. The ratio between
evapotranspiration and precipitation (evapotranspiration index = ET_idx) is a very useful indicator
how precipitation is used by plants during photosynthesis. Calculate the ET_idx for all sites and
visualise it. What surprised you and how do the ET_idx differs across ecosystems? Discuss.*


```{r}
sites$et_idx <- sites$et/sites$prec

ggplot(sites,
       aes(site, et_idx, colour = site))+
         geom_boxplot()+
         labs(title="Evapotranspiration index", x="Different site", y="Evapotranspiration index")
        
```

As we can see,evapotranspiration index is the fraction between precipitation and what is coming back to atmosphere through evapotranspiration. 

what is intercepted by cannopy and what is being lose by soil hydrology processes. Other variables such as wind speed, surface humidity and solar radiation are not being taken into account in the calculation of this index, which probably are the ones causing differences on what it could have been expected at the beginning for some of the places. 

we can see a higher evapotranspiration index in places where there is plenty precipitation and evapotranspiration rate by canopy. Hainich and Hyytiala are represented as the higher, this can be due to their higher latitudes providing this ecosystems with more precipitation rate. Guyaflux is at lower latitude and thus more close to the equator. This means a more tropical weather where high amount of rain and sun is normally occurring. Collelongo is place in a drier climate with high radiation but less annual precipitation rates. 





