---
title: '5th Protocol: Precipitation'
author: "Cheyenne Rueda and Simone Massaro"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    includes:
      in_header: "../preamble.tex"
  html_document:
    df_print: paged
header-includes:
  - \fancyhead[L]{Shortwave radiation}
---

## Analysis
```{r, include = F}
library(tidyverse)
library(patchwork)
```
```{r, message=F}
prec <- read_csv("../Data_lectures/5_Precipitation/P_4sites.csv")
et <- read_csv("../Data_lectures/5_Precipitation/ET_4sites.csv")
sites <- read_csv("station_data.csv") %>%
  rename(site=`Site-abb`, full_name = Site)
```
```{r}
prec_avg <- prec %>%
  select(-Date) %>%
  gather("site", "prec") %>%
    group_by(site) %>%
  summarise(prec=mean(prec))

et_avg <- et %>%
  select(-Date) %>%
  gather("site", "et") %>%
    group_by(site) %>%
  summarise(et=mean(et))

sites_avg <- sites %>%
  inner_join(et_avg, by="site") %>%
  inner_join(prec_avg, by="site")
```

## ET and precipitation at different latitudes

*Visualise the annual sums of evapotranspiration and precipitation according to its geographic
latitude (boxplot?). How and why does precipitation and evapotranspiration varies? Which impact
has the underlying ecosystem?*

```{r}
# this t makes no sense as name TODO chanve this
prec_t <- prec %>%
  gather("site", "prec", -Date)

et_t <- et %>%
  gather("site", "et", -Date) 

prec_et <- left_join(prec_t, et_t, by=c("Date", "site"))

sites <- sites %>%
  inner_join(prec_et, by="site") 
```

```{r}
sites
```


```{r}
(prec_box <-  ggplot(sites, aes(Lat, prec, colour=site)) +
  geom_boxplot() +
   labs(title="Cumlative precipitation at different latitudes",
        y="Total yearly precipitation (mm)",
        x="Latitude (deg)"))
```
```{r}
(et_box <- ggplot(sites, aes(Lat, et, colour=site)) +
  geom_boxplot() +
    labs(title="Cumlative evapotranspiration at different latitudes",
        y="Total yearly evapotranspiratio (mm)",
        x="Latitude (deg)"))
```
## Evapotranspiration index

*Precipitation is used as a driver/source for evapotranspiration. The ratio between
evapotranspiration and precipitation (evapotranspiration index = ET_idx) is a very useful indicator
how precipitation is used by plants during photosynthesis. Calculate the ET_idx for all sites and
visualise it. What surprised you and how do the ET_idx differs across ecosystems? Discuss.*



