---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(lubridate)
```


Plot air temperature inside the canopy and above the canopy.
Indicate time periods when the two temperatures differ and explain
why.

```{r}
df <- read_csv("../Data_lectures/3_Air_temperature/Hainich_T_air_soil_degC.csv")
```



```{r}
ggplot(df, aes(x=Date)) +
  geom_line(aes(y=TA_44m, colour="44m")) +
  geom_line(aes(y=TA_2m, colour="2m"))
```
```{r}
df$diff_t <- df$TA_44m - df$TA_2m
```

```{r}
ggplot(df, aes(x=Date)) +
  geom_line(aes(y=diff_t, colour="diff 44 2 m")) 
```
```{r}
df %>%
  filter(Date > as_date("2019-06-1") & Date < as_date("2019-07-1"))  %>%
  ggplot(aes(x=Date)) +
  geom_line(aes(y=diff_t, colour="diff 44 2 m")) 
```

```{r}
df %>%
  filter(Date > as_date("2019-06-1") & Date < as_date("2019-06-8"))  %>%
  ggplot(aes(x=Date)) +
  geom_line(aes(y=diff_t, colour="diff 44 2 m")) 
```
Plot air temperature in 2 m height and soil temperature in 0.02 m
depth. Explain the characteristics of the time series?

Estimate the response time of a mercury thermometer (soil
temperature) and an Assmann thermometer (air temperature).
Visualize and discuss differences.
