---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(lubridate)

between_dates <- function(x, start, end){
  between(x, 
          parse_date_time(start, c("dm", "dmy")),
          parse_date_time(end, c("dm", "dmy")))
}
```


Plot air temperature inside the canopy and above the canopy.
Indicate time periods when the two temperatures differ and explain
why.

```{r}
temp <- read_csv("../Data_lectures/3_Air_temperature/Hainich_T_air_soil_degC.csv") %>% 
  mutate(diff_canopy = TA_44m - TA_2m) %>% 
  drop_na()
```

```{r}
temp_d <- temp %>% 
  group_by(Date = round_date(Date, "1d")) %>% 
  summarize_all(mean)
```


```{r}
temp %>% 
  filter(Date > as_date("2019-06-1") & Date < as_date("2019-06-7"))  %>%
  ggplot(aes(x=Date)) +
  geom_line(aes(y=TA_44m, colour="44m")) +
  geom_line(aes(y=TA_2m, colour="2m"))
```

```{r}
ggplot(temp, aes(x=Date)) +
  geom_line(aes(y=TA_44m, colour="44m")) +
  geom_line(aes(y=TA_2m, colour="2m"))
```


```{r}
temp_d %>% 
  ggplot(aes(Date, temp, col = height)) +
  geom_line(aes(y=TA_44m, colour="44m")) +
  geom_line(aes(y=TA_2m, colour="2m"))
```
```{r}
temp %>% 
  group_by(Date = round_date(Date, "6h")) %>% 
  summarize_all(mean) %>% 
  ggplot(aes(Date)) +
  geom_line(aes(y=diff_canopy, colour="diff"))
```
```{r}
temp %>% 
  filter(month(Date) == 4) %>% 
  group_by(Date = round_date(Date, "30min")) %>% 
  summarize_all(mean) %>% 
  pivot_longer(c(-Date, -Tsoil_002m_degC), names_to = "h", values_to = "temp") %>% 
  ggplot(aes(Date, temp, col = h)) +
  geom_line()
```
```{r}
temp %>% 
  filter(between_dates(Date, "15 April 2019", "")) %>% 
  group_by(Date = round_date(Date, "30min")) %>% 
  summarize_all(mean) %>% 
  pivot_longer(c(TA_2m, TA_44m, diff_canopy), names_to = "h", values_to = "temp") %>% 
  ggplot(aes(Date, temp, col = h)) +
  geom_line()
```



```{r}
temp %>% 
  pivot_longer(c(-Date, -Tsoil_002m_degC), names_to = "h", values_to = "temp") %>% 
  ggplot(aes(h, temp, col = h)) +
  geom_boxplot()
```




```{r}
temp$diff_t <- temp$TA_44m - temp$TA_2m
```

```{r}
ggplot(temp, aes(x=Date)) +
  geom_line(aes(y=diff_t, colour="diff 44 2 m")) 
```
```{r}
temp %>%
  filter(Date > as_date("2019-06-1") & Date < as_date("2019-07-1"))  %>%
  ggplot(aes(x=Date)) +
  geom_line(aes(y=diff_t, colour="diff 44 2 m")) 
```

```{r}
temp %>%
  filter(Date > as_date("2019-06-1") & Date < as_date("2019-06-8"))  %>%
  ggplot(aes(x=Date)) +
  geom_line(aes(y=diff_t, colour="diff 44 2 m")) 
```


Plot air temperature in 2 m height and soil temperature in 0.02 m
depth. Explain the characteristics of the time series?


```{r}
temp %>% 
  #filter(between_dates(Date, "15 April 2019", "")) %>% 
  group_by(Date = round_date(Date, "1d")) %>% 
  summarize_all(mean) %>% 
  pivot_longer(c(TA_2m, Tsoil_002m_degC), names_to = "h", values_to = "temp") %>% 
  ggplot(aes(Date, temp, col = h)) +
  geom_line()
```

```{r}
temp %>% 
  filter(between_dates(Date, "1 May 2019", "30 May 2019")) %>% 
  group_by(Date = round_date(Date, "1h")) %>% 
  summarize_all(mean) %>% 
  pivot_longer(c(TA_2m, Tsoil_002m_degC), names_to = "h", values_to = "temp") %>% 
  ggplot(aes(Date, temp, col = h)) +
  geom_line()
```

```{r}
temp %>% 
  group_by(Date = round_date(Date, "1d")) %>% 
  summarize_all(mean) %>% 
  ggplot(aes(TA_2m, Tsoil_002m_degC)) +
  geom_point() +
  geom_smooth(method = "lm", se=F)
```



Estimate the response time of a mercury thermometer (soil
temperature) and an Assmann thermometer (air temperature).
Visualize and discuss differences.
