---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
```


## Resistance sensor 

```{r}

res_wire <- 3.7 # Omega. This has been measured in the field
r_0 <- 100 # Omega at 100 degrees
a <- 4e-3


#' converts resistance to temperature
get_temp <- function(resistance) {
  (((resistance - res_wire) / 100) - 1) / 4e-3 
}
```

we heat up the sensor and then measure every 10 seconds the temperature while it cooled down


load data
```{r}
temp_res <- read.csv("resistance_sensor_cooling response.csv", header = T)
```

add the time (in seconds) and the temperature after conversion from resistance
```{r}
temp_res <- temp_res %>%
  mutate(time = 0:(nrow(temp_res)-1) * 10,
         temp = get_temp(resistance) )
```


To estimate the response time we can start from the following equation that describes the temperature decrease over time

$$T_t = T_a + (T_0 - T_a) e^{\frac{T} {-\tau}}$$
by inverting this formula and fitting a linear model we estimate $\tau$

```{r}
T_a <- 13.0 # calculations done in the field
T_0 <- temp_res$temp[1] # first measure 

temp_res <- mutate(temp_res, log_t = log((T_0 - T_a) / (temp - T_a)) )
```

```{r}
ggplot(temp_res, aes(x=time, y=log_t)) +
  geom_line()
```

```{r}
(model_res <- lm(time ~ log_t, data=temp_res))
```

The time response rate for the resistance sensor is `r round(coef(model_res)[1])` seconds.

## Mercury termometer

```{r}
temp_merc <- read.csv("mercury_termometer_cooling_response.csv", header = T)
```


```{r}
T_a <- 7.0 # ambient temperature
T_0 <- temp_merc$temp[1] # first measure 

temp_merc <- mutate(temp_merc, log_t = log((T_0 - T_a) / (temp - T_a)) )
```


```{r}
ggplot(temp_merc, aes(x=time, y=log_t)) +
  geom_line()
```

```{r}
(model_merc <- lm(time ~ log_t, data=temp_merc))
```

The time response rate for the resistance sensor is `r round(coef(model_res)[1])` seconds.



