---
title: '6th Protocol: Air Pressure'
author: "Cheyenne Rueda and Simone Massaro"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: yes
    number_sections: yes
    includes:
      in_header: ../preamble.tex
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    number_sections: yes
  pdf_document:
    toc: yes
bibliography: ../bioclimatology-references.bib
header-includes: \fancyhead[L]{Air Pressure}
editor_options:
  markdown:
    wrap: 72
---

\newpage

# Motivation

Air pressure is commonly known by experts as a factor that is constantly
regulating the characteristics of natural systems. Changes in cloud
densities, clouds movement and humid conditions of air. All these are
examples of how air dynamic alters natural components of earth. But,
what is the reason that explains the movement of air pressure? The
molecules floating in the air have different directional movements or
for instance, changes in its momentum. Following the same process as
water particles when heated, air molecules move faster when air is hot.
Air pressure is as well strongly related with the weather humidity
conditions. Air is fluid. The molecules can easily move in all
directions [@bonan_climate_2019].

Theory explains the different layers the atmosphere is composed by. Each
of these layers presents air in different amounts. The air of the
atmosphere exerts pressure through all of them and this is known as *air
pressure* or *atmospheric pressure*, which makes force towards the
center of the planet from all directions, to all the earth component.
More concretely, the troposphere (the coolest layer to earth surface)
has the higher amount of air.

# Background

There are different processes that affects air pressure. The main one is
the reduction of air pressure with the increase of elevation. The
pressure depends on the weight of the air column above, at a high
elevation there is less air above and thus, the pressure is lower.

This dependency of pressure with elevation is commonly used in
altimeters to estimate the elevation. However, in case of weather
measurements the height effects needs to be removed.

The air pressure has an exponential decay with height. The following
formula can be used to estimate the pressure using the measured pressure
at a know elevation.

$$p(0) = p(z) exp({\frac{g \Delta z} {R_d T}})$$

The air temperature is a important component of the formula, as warm air
is less dense and therefore the air column weights less. However, air
temperature also depends on altitude, hence, the mean temperature over
the column is used considering an estimate decrease in temperature of
$-0.65 K /100m$. The following equation can be used

$$T=T_{station}+0.00325*z$$

Another correction can be made for air humidity, as wet air is less
dense than dry air. This is done by estimating the virtual temperature,
which corresponds to the temperature where dry air would have the same
density of the wet air. The virtual temperature is always higher than
the real one. The following equation can be used for the correction

$$T_v = T (1 + 0.608q)$$

where $q$ is the specific humidity in ($Kg/Kg$)

The inverse of the previous formula can be used to calculate the
difference of height between 2 points with known pressures.

$$\Delta z = -\frac{R_d}{g T} log(\frac{p_z} {p_0})$$

One peculiarity of air pressure is the wide range of units used around
the globe to measure it. The SI defines the Pa ($N/m^2$) as the unit for
pressure. However, this is a small value so hPa (100 Pa) is commonly
used as a reference amount of this unit. Another unit that commonly used
in barometers is the torr or mmHg, that originates from the millimeters
of mercury used in the first barometers. Those are the values to change
between units: 760 mmHg = 760 torr = 1013.25 hPa.

Pressure influences the boiling point temperature of water. The
following equation can be used to estimate the relationship:

$$T_{boil} = 100 + 2.804 \times 10^{-2} (p-1013.25hPa) - 1.384 \times 10^{-5} (p-1013.25hPa)^2$$

# Sensors and measuring principle

There are several sensors to measure the air pressure and each of these
use different measurement principles.

-   **Mercury barometer**. This is the oldest barometer and works by
    having a column of mercury in a tube with vacuum on one side and air
    in the other. On the mercury there is the gravitational force that
    make it going down, while the air pressure pushes the column up.
    This two forces reaches and equilibrium and therefore it is possible
    to read the pressure using the height of the mercury column. This
    sensor is not commonly used nowadays anymore. First of all, because
    mercury is dangerous and then, it also requires error corrections
    for both: temperature (mercury expands with higher temperatures) and
    gravity acceleration constant, which changes depending on altitude
    and latitude.

-   **Aneroid barometer**. They have an aneroid capsule with vacuum (or
    low pressure) inside, air pressure tends to reduce the collapse the
    capsule while a spring keeps in open. By measuring the width of the
    capsule is possible to estimate the air pressure. The width of the
    capsule can be measured both in analog instruments or digital one,
    using a capacitor. Those are the most widely used pressure sensors
    as they are compact, reliable and require no error correction.

-   **Boiling barometer**. First, it measures the boiling temperature of
    water, and then, uses this information to estimate the air pressure.
    There is a heater to make water boil and then an accurate
    thermometer measure the temperature of the water vapour. The main
    disadvantage is their reduced convenience due to the procedure to
    boil water at each sample, but they can have a high accuracy, up to
    0.5 hPa [@richner_water_1996].

\newpage

# Analysis

```{r, include=F}
library(tidyverse)
theme_set(theme_bw())
```

```{r, message = F}
pres <- read_csv("../Data_lectures/6_Air_pressure/TA_RH_PA_Leinefelde.csv")
```

```{r}
#utility funcs from air humidity notebook

# temp is in degrees celcius
get_es <-  function(ta)  6.1078 * exp((17.08085 * ta) / (234.175 + ta))

rh2ea <- function(rh, e_s) rh/100 * e_s
get_spec_hum <- function(e_a, p) 0.622 * e_a / (p - 0.378 * e_a) # note Kg Kg-1

c2k <- function(c) c + 273.15
k2c <- function(k) k - 273.15

# get virtual temperature. Ta is the air temp and q the specific humidity
get_tv <- function(ta, q) ta * (1 + 0.608 * q)

Rd <- 287.05 # J Kg-1 K-1 gas constant of dry air
get_press_sea_level <- function(pz, tv, Dz, g = 9.81) {
  pz * exp((g * Dz)/(Rd * c2k(tv)))
}

```

## Air pressure sea level

*Calculate the air pressure at sea level for the Leinefelde site.*

```{r}
heigth_diff <- 451 + 44 # elevation + tower height

pres <- pres %>%
  mutate(
    es = get_es(TA_degC),
    ea = rh2ea(RH_Perc, TA_degC),
    q = get_spec_hum(ea, PA_hPa),
    tv = get_tv(TA_degC, q),
    p0 = get_press_sea_level(PA_hPa, tv, heigth_diff)
  )
```

\newpage

```{r, "air-pressure-plot", fig.cap="Air pressure at Leinefelde and estimated air pressure at sea level. Pressure has been corrected for air humidity. Data from Leinefelde flux tower (451 m + 41m tower) July 2020 - May 2021, 30 min frequency."}
pres %>%
        gather("location", "pressure", p0, PA_hPa, factor_key = T) %>%
ggplot(aes(Date, pressure, color=location)) +
  geom_line() +
labs(y="Pressure (hPa)") +
scale_colour_discrete(name="Location", labels = c("Sea level", "Leinefelde"))
```

The air pressure at sea level is always higher than at Leinefelde
(Figure \@ref(fig:air-pressure-plot)) The difference is around 53 hPa,
and it is quite constant during the year. In the plot it is also
possible to see that there are no clear seasonal patterns during the
year as the pressure oscillate roughly +/- 25 hPa around the mean.
Moreover, it is also interesting to notice the stability of the air
pressure in a short time frame, as the data plotted has a 30 min
frequency but no high frequency patterns can be observed.

\newpage

## Air pressure Brocken and water boling temperature

*Calculate the air pressure at top of the Brocken mountain (1141 m) for
a pressure of 991.3 hPa and an air temperature of 15°C at the North
campus (185 m). Assume a mean air temperature decrease of 0.5 K/100 m.
At which temperature would water boil at the North Campus and on the
Brocken?*

```{r}
pa_nc <- 991.3  # Air pressure north campus
Rd <- 287.05 # J Kg-1 K-1 gas constant of dry air
g <- 9.81# m/s^2
z1 <- 1141 #m Brocken mountain Height
z0 <- 185 #m North campus Height
Z <- z0-z1
ta <- 15 # C measuered at the north campus
t_column <- ta + 0.00325 * Z #correction of temperature for elevation
#hPa would be the pressure at the Brocken mountain
pa_brocken<-  (991.3 * exp((g * Z)/(Rd * c2k(t_column)))) %>% round(2)

```

On the top of the Brocken the air pressure would be `r pa_brocken` hPa

```{r}
t_boil_nc <- (100 + 2.804e-2 * (pa_nc-1013.25) - 1.384e-5 *(pa_nc-1013.25)^2 ) %>%
                round(2)
t_boil_brocken <- (100 + 2.804e-2 * (pa_brocken-1013.25) - 1.384e-5 *
                     (pa_brocken-1013.25)^2 ) %>% round(2)
```

The water would boil at `r t_boil_nc` °C at the North Campus and
`r t_boil_brocken` °C on the Brocken. As expected, the boiling
temperature is lower at a higher elevation.

The pressure value of the Brocken mountain was calculated using the
formula mentioned in the theory background, adding correction for
temperature decrease. $$p(0) = p(z) exp({\frac{g \Delta z} {R_d T}})$$

# References

