---
title: "9th Protocol: Evapotranspiration"
author: "Cheyenne Rueda and Simone Massaro"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: true
    includes:
      in_header: "../preamble.tex"
  html_notebook:
    number_sections: true
header-includes:
  - \fancyhead[L]{Evapotranspiration}
editor_options: 
  markdown: 
    wrap: 72
---

\newpage
# Evapotranspiration

## Motivation

In this chapter we will focus on evapotranspiration. This term is well
known in many environmental study fields, but, what is the most precise
definition given to this term?

Evapotranspiration (ET) is referred according to the relationship of two
other factors, such as the soil water evaporation and plant
transpiration. The actual evapotranspiration is the one that is measured
from these two factors mentioned before, considering changes and not
ideal conditions in the meteorological conditions and the type of soil
(Labedzki, 2011). In contrast, potential evapotranspiration is that one
that occurs with ideal conditions such as abundance of water storage and
meteorological conditions. Some differences between potantial and actual
evapotranspiration are the infiltration capacity of soil, possible
diseases, pH of soil and its fertility.

Labedzki, L. (Ed.). (2011). Evapotranspiration. BoD--Books on Demand.

## Background


## Sensors and measuring principle


There are many different ways to measure evapotranspiration,some are
more accessible than others and depending the scenario some of them
might be more suitable than others. The measurement of
evapotranspiration claims quantitative data, and this data can be
measure by water evaporation and the energy flux between soil and
atmosphere.Rana and Kater (2000), describes the different ways to
measure evapotranspiration based on hydrological, micrometeorological,
plant physiology and analytical approaches. The first one includes: soil
water balance and weighing lysimeters, the second one, energy balance
and Bowen ration, aerodynamic method and eddy covariance. Plant
physiology approach is based on sap flow method or chambers system. Last
one, analytical approach are based on Penman-Montheith model. After this
one, empirical approach can also be taken into account, such as process
based on crop coefficient approach and soil water balance modeling
[@articlenoauthor_rana_nodate].

Some of the most frequently measurements applied to evapotranspiration
and that we have seeing in class are:

Evaporation pan: a circle panel where precipitation is accumulated and
then, with the help of a measuring bucket and based on a scale, water
los can be measured by the difference between the potential
evapotranspiration of 2 days. Some of the erros are the expansion of
water, wrong reading, limited recording for a volume of water and the
possibility of the oasis effect to occur.

Piché evaporimeter

Weighing lysimeter

Bowen ratio energy balance method

## Analysis


### Potential evapotranspiration


```{r}
library(tidyverse)
library(lubridate)
library(scales)
theme_set(theme_bw())
et <- read_csv(here::here("Data_lectures/09_Turbulent_fluxes_I_ET/ET_data_forst_botanical_garden.csv") , locale = locale(decimal_mark = ",")) %>% 
  rename(loc_id = replicates)
meteo <- read_csv(here::here("Data_lectures/09_Turbulent_fluxes_I_ET/MeteoData_BotanicalGarden.csv"))
```

```{r}
et
```

```{r}
meteo
```

```{r}
#' Potential evapotranspiration using Priestley-Taylor equation
calc_pet <- function(T_air, Rn, G){
  g <- 0.067 # kPa K -1
  s <- ( 4098 * (0.6108 * exp((17.27 * T_air ) / (T_air +237.3) ) ) ) / ( T_air + 237.3 )^2
  pet <- 1.26 * s * (Rn - G)/ (s + g)
}

```

```{r}
#adding R_n and G to the 
meteo_d <- meteo %>% 
  # Need to convert from W (J/s) to MJ/d, using a factor 0.0864
  # calculating with high frequency data and then averaging over the day
  mutate(PET = calc_pet(TA_degC, `NetRadiation_Wm-2`, `GroundHeatflux_Wm-2`)) %>% 
  group_by(Date = floor_date(Date, "day")) %>% 
  summarise(PET = mean(PET))
```

```{r, pet, fig.cap="Potential evapotranspiration estimated using the Priestley-Taylor equation. Data from the botanical garden for the 25-29th of June 2021."}
(meteo_d <- meteo %>% 
  group_by(Date = floor_date(Date, "day")) %>% 
  # Need to convert from W (J/s) to MJ/d, using a factor 0.0864
  summarise(R_n_d = mean(`NetRadiation_Wm-2`) * 0.0864, G_d = mean(`GroundHeatflux_Wm-2`) * 0.0864, T_air = mean(TA_degC)) %>% 
  mutate(
    PET = calc_pet(T_air, R_n_d, G_d)
  ))
```

```{r}
ggplot(meteo_d, aes(Date, PET)) +
  geom_line() +
  labs(y="PET (mm)")
```

During the four days of measurement the potential ET has been estimated
using the Priestley-Taylor (Figure \@ref(fig:pet)). The daily potential
ET varies a significantly during the 4 days, ranging from 6mm to almost
12 mm. This reflects the change in the weather conditions, as the day 1
and 4 were cloudy and colder.

```{r}
et <- et %>% 
  group_by(loc_id) %>% 
  mutate(
    et_pan = lag(pan_height_mm) - pan_height_mm,
  )
```

TODO

### Piché evaporimeters

The potential ET has been measured using Piché evaporimeters in
different locations around the botanical garden. The measures values
have been compared between the different locations and the evaporation
pan

```{r}
pich_d <- 3 #cm
pich_inn_d <- 0.9 # cm
# calc area exposed to air:
# 2 times the area of the pare dish (two sides) - the area of glass
pich_dish_area <- 2 * (pi / 4 * pich_d ^ 2) - (pi / 4 * pich_inn_d ^ 2) # cm^2
# area inside the tube
pich_int_area <- (pi/4 * pich_inn_d ^2 ) # cm^2
```

```{r}
et <- et %>% 
  group_by(loc_id) %>% 
  mutate(
    # here the scale is the opposite, the lower the number the more the water
    diff_pich = pich_height_cm - lag(pich_height_cm),
    # need to convert to mm of ET
    et_pich = diff_pich * pich_dish_area * pich_int_area / 10
  )
```

```{r, pan-pich, fig.cap="Comparison between evapotranspiration pan and the "}
spot_cols <- hue_pal()(4)
names(spot_cols) <- unique(et$spot)
et %>% 
  drop_na() %>% #removing first empty day
ggplot(aes(date)) +
  geom_line(aes(y=et_pan, col="Pan"), size=.7) +
  geom_jitter(aes(y = et_pich, col=spot), width = .06, height = 0) +
  scale_color_manual(values=c(spot_cols, "Pan" = "black" )) +
  labs(y="Potential evapotranspiration (mm)", colour="Location")
```

```{r}
et %>% 
  drop_na() %>% 
ggplot(aes(spot, et_pich, col=spot)) +
  geom_boxplot()
```

```{r}
et %>% 
  left_join(meteo_d, by = c("date"= "Date")) %>% 
  gather("type", "et", et_pan, PET) %>% 
  ggplot(aes(date, et, col=type)) +
  geom_line() +
  labs(y= "Evapotranspiration (mm)")
```

```{r}
et
```

