---
title: "9th Protocol: Evapotranspiration"
author: "Cheyenne Rueda and Simone Massaro"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: true
    includes:
      in_header: "../preamble.tex"
  html_notebook:
    number_sections: true
header-includes:
  - \fancyhead[L]{Evapotranspiration}
editor_options: 
  markdown: 
    wrap: 72
---

\newpage

# Evapotranspiration

## Motivation

In this chapter we will focus on evapotranspiration. Evapotranspiration
(ET) is referred according to the relationship of two other factors,
such as the soil water evaporation and plant transpiration. The actual
evapotranspiration is the one that is measured from these two factors
mentioned before, considering changes and not ideal meteorological
conditions and type of soil (Labedzki, 2011). In contrast, potential
evapotranspiration occurs with ideal conditions such as abundance of
water storage and meteorological conditions. Some differences between
potantial and actual evapotranspiration are the infiltration capacity of
soil, possible diseases, pH of soil and its fertility.

Labedzki, L. (Ed.). (2011). Evapotranspiration. BoD--Books on Demand.

## Background

## Sensors and measuring principle

There are many different ways to measure evapotranspiration,some are
more accessible than others and depending the scenario some of them
might be more suitable than others. The measurement of
evapotranspiration claims quantitative data, and this data can be
measure by water evaporation and the energy flux between soil and
atmosphere.Rana and Kater (2000), describes the different ways to
measure evapotranspiration based on hydrological, micrometeorological,
plant physiology and analytical approaches. The first one includes: soil
water balance and weighing lysimeters, the second one, energy balance
and Bowen ration, aerodynamic method and eddy covariance. Plant
physiology approach is based on sap flow method or chambers system. Last
one, analytical approach are based on Penman-Montheith model. After this
one, empirical approach can also be taken into account, such as process
based on crop coefficient approach and soil water balance modeling
[@articlenoauthor_rana_nodate].

Some of the most frequently measurements applied to evapotranspiration
and that we have seeing in class are:

Evaporation pan: a circle panel where precipitation is accumulated and
then, with the help of a measuring bucket and based on a scale, water
loss can be measured by the difference between the potential
evapotranspiration of 2 days. Some of the errors are related to the
expansion of water, wrong reading, limited recording for a volume of
water and the possibility of the oasis effect to occur.

Piché evaporimeter: is a type of atometer applied in the measurement
rate of evaporation from a wet disc with absorbent paper. Results are
dependent on wind speed that goes through the disc, as well as the wet
bulb saturation deficit. The rate of evaporation is usually expressed as
the volume of water evaporated per unit area in unit time.

Weighing lysimeter: they allow the mass or volumetric soil water content
variation to be measured by weighing the lysimeter and determining its
change of mass over time. They can measure the net infiltration from
precipitation or irrigation systems and the quantity of net evaporation
between different wetting events (Meissner et al., 2020).

{Meissner, R., Rupp, H., & Haselow,
L. (2020). Use of lysimeters for monitoring soil water balance
parameters and nutrient leaching. In *Climate change and soil
interactions* (pp. 171-205). Elsevier.}

Bowen ratio energy balance method: also known as BREB method has been
widely used to quantify water balance. It estimates the latent heat flux
from a surface using measurements of air temperature and humidity
gradients, net radiation and soil heat flux. In comparison with other
methods such as eddy covariance or weighting lysimeters, is an indirect
method (Todd et al., 200). Some advantages are that it does not require
information about aerodynamic characteristics of zone of interest, it
integrates latent hear fluxes over a wide areas and gives an estimation
of these flux in short time period intervals (e.g. less than half hour).

{Todd, R. W., Evett, S. R., &
Howell, T. A. (2000). The Bowen ratio-energy balance method for
estimating latent heat flux of irrigated alfalfa evaluated in a
semi-arid, advective environment. *Agricultural and Forest Meteorology*,
*103*(4), 335-348.}

## Analysis

### Potential evapotranspiration

```{r, warning=FALSE}
library(tidyverse)
library(lubridate)
library(scales)
theme_set(theme_bw())

et <- "Data_lectures/09_Turbulent_fluxes_I_ET/ET_data_forst_botanical_garden.csv" %>% 
  here::here() %>% 
  read_csv(locale = locale(decimal_mark = ",")) %>% 
  rename(loc_id = replicates)
meteo <- 
  "Data_lectures/09_Turbulent_fluxes_I_ET/MeteoData_BotanicalGarden.csv" %>% 
  here::here() %>% 
  read_csv()
```

```{r}
#' Potential evapotranspiration using Priestley-Taylor equation
calc_pet <- function(T_air, Rn, G){
  g <- 0.067 # kPa K -1
  s <- ( 4098 * (0.6108 * exp((17.27 * T_air ) / (T_air +237.3) ) ) ) / ( T_air + 237.3 )^2
  pet <- 1.26 * s * (Rn - G)/ (s + g)
}

```

```{r}
#adding R_n and G to the 
meteo_d <- meteo %>% 
  # Need to convert from W (J/s) to MJ/d, using a factor 0.0864
  # calculating with high frequency data and then averaging over the day
  mutate(PET = calc_pet(TA_degC, `NetRadiation_Wm-2`, `GroundHeatflux_Wm-2`)) %>% 
  group_by(Date = floor_date(Date, "day")) %>% 
  summarise(PET = mean(PET))
```

```{r, pet, fig.cap="Potential evapotranspiration estimated using the Priestley-Taylor equation. Data from the botanical garden for the 25-29th of June 2021."}
(meteo_d <- meteo %>% 
  group_by(Date = floor_date(Date, "day")) %>% 
  # Need to convert from W (J/s) to MJ/d, using a factor 0.0864
  summarise(R_n_d = mean(`NetRadiation_Wm-2`) * 0.0864,
            G_d = mean(`GroundHeatflux_Wm-2`) * 0.0864,
            T_air = mean(TA_degC)) %>% 
  mutate(
    PET = calc_pet(T_air, R_n_d, G_d)
  ))
```

```{r}
ggplot(meteo_d, aes(Date, PET)) +
  geom_line() +
  labs(y="PET (mm)")
```

During the four days of measurement the potential ET has been estimated
using the Priestley-Taylor (Figure \@ref(fig:pet)). The daily potential
ET varies a significantly during the 4 days, ranging from 6mm to almost
12 mm. This reflects the change in the weather conditions, as the day 1
and 4 were cloudy and colder.

```{r}
et <- et %>% 
  group_by(loc_id) %>% 
  mutate(
    et_pan = lag(pan_height_mm) - pan_height_mm,
  )
```

TODO

### Piché evaporimeters

The potential ET has been measured using Piché evaporimeters in
different locations around the botanical garden. The measures values
have been compared between the different locations and the evaporation
pan

```{r}
pich_d <- 3 #cm
pich_inn_d <- 0.9 # cm
# calc area exposed to air:
# 2 times the area of the pare dish (two sides) - the area of glass
pich_dish_area <- 2 * (pi / 4 * pich_d ^ 2) - (pi / 4 * pich_inn_d ^ 2) # cm^2
# area inside the tube
pich_int_area <- (pi/4 * pich_inn_d ^2 ) # cm^2
```

```{r}
et <- et %>% 
  group_by(loc_id) %>% 
  mutate(
    # here the scale is the opposite, the lower the number the more the water
    diff_pich = pich_height_cm - lag(pich_height_cm),
    # need to convert to mm of ET
    et_pich = diff_pich * pich_dish_area * pich_int_area / 10
  )
```

```{r, pan-pich, fig.cap="Comparison between evapotranspiration pan and the "}
spot_cols <- hue_pal()(4)
names(spot_cols) <- unique(et$spot)
et %>% 
  drop_na() %>% #removing first empty day
ggplot(aes(date)) +
  geom_line(aes(y=et_pan, col="Pan"), size=.7) +
  geom_jitter(aes(y = et_pich, col=spot), width = .06, height = 0) +
  scale_color_manual(values=c(spot_cols, "Pan" = "black" )) +
  labs(y="Potential evapotranspiration (mm)", colour="Location") +
  scale_color_colorblind()
```

```{r, fig.cap="Piche evaporimeters"}
et %>% 
  drop_na() %>% 
ggplot(aes(spot, et_pich, col=spot)) +
  geom_boxplot() +
  scale_color_colorblind()
```

```{r, fig.cap="Pan vs potential evapo"}
et %>% 
  left_join(meteo_d, by = c("date"= "Date")) %>% 
  gather("type", "et", et_pan, PET) %>% 
  ggplot(aes(date, et, col=type)) +
  geom_line() +
  labs(y= "Evapotranspiration (mm)") +
  scale_color_colorblind()
```
